---
title: "Running calculate_aci() End to End"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running calculate_aci() End to End}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(xaci)
```

`calculate_aci()` is the main entry point of the package: it computes all
six components (see `vignette("xaci-components")`) and combines them into
the index itself, in a single call. This vignette runs it end to end on the
same kind of synthetic dataset used elsewhere in this package's vignettes,
and explains the shape of its output.

## Building the synthetic dataset

```{r}
build_synthetic_nc <- function(path, var, unit, lon, lat, time_vec, origin, vals) {
  time_hours <- as.numeric(difftime(time_vec, origin, units = "hours"))
  dim_lon  <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat  <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  dim_time <- ncdf4::ncdim_def(
    "time", paste0("hours since ", format(origin, "%Y-%m-%d %H:%M:%S")),
    time_hours, unlim = TRUE
  )
  ncvar <- ncdf4::ncvar_def(var, unit, list(dim_lon, dim_lat, dim_time),
                            missval = NA, prec = "double")
  nc <- ncdf4::nc_create(path, list(ncvar))
  ncdf4::ncvar_put(nc, ncvar, vals)
  ncdf4::nc_close(nc)
  invisible(path)
}

build_synthetic_mask <- function(path, lon, lat) {
  dim_lon <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  var_mask <- ncdf4::ncvar_def("country", "1", list(dim_lon, dim_lat),
                               missval = NA, prec = "double")
  nc <- ncdf4::nc_create(path, list(var_mask))
  ncdf4::ncvar_put(nc, var_mask, matrix(1, length(lon), length(lat)))
  ncdf4::nc_close(nc)
  invisible(path)
}

set.seed(7)
lon      <- c(-1, 0, 1)
lat      <- c(43, 44, 45)
origin   <- as.POSIXct("1900-01-01 00:00:00", tz = "UTC")
time_vec <- seq(as.POSIXct("2011-01-01 00:00", tz = "UTC"),
                as.POSIXct("2013-12-31 23:00", tz = "UTC"), by = "hour")
nlo <- length(lon); nla <- length(lat); nt <- length(time_vec)

# A mild warming/drying trend, so 2013 (the one out-of-reference year, see
# below) shows genuine anomalies rather than pure noise.
trend <- seq_len(nt) / nt

seasonal_t <- 288 + 10 * sin(2 * pi * seq_len(nt) / (24 * 365)) + 0.6 * trend
t2m_vals <- array(NA_real_, c(nlo, nla, nt))
for (i in seq_len(nlo)) for (j in seq_len(nla))
  t2m_vals[i, j, ] <- seasonal_t + (i + j) + rnorm(nt, sd = 1.5)

tp_vals <- array(0, c(nlo, nla, nt))
for (i in seq_len(nlo)) for (j in seq_len(nla)) {
  rain_hours <- rbinom(nt, 1, 0.08 * (1 - 0.3 * trend))
  tp_vals[i, j, ] <- rain_hours * rexp(nt, rate = 800)
}

u10_vals <- array(rnorm(nlo * nla * nt, mean = 3, sd = 4), c(nlo, nla, nt))
v10_vals <- array(rnorm(nlo * nla * nt, mean = 1, sd = 4), c(nlo, nla, nt))

t2m_file  <- tempfile(fileext = ".nc")
tp_file   <- tempfile(fileext = ".nc")
u10_file  <- tempfile(fileext = ".nc")
v10_file  <- tempfile(fileext = ".nc")
mask_file <- tempfile(fileext = ".nc")

build_synthetic_nc(t2m_file, "t2m", "K",     lon, lat, time_vec, origin, t2m_vals)
build_synthetic_nc(tp_file,  "tp",  "m",     lon, lat, time_vec, origin, tp_vals)
build_synthetic_nc(u10_file, "u10", "m s-1", lon, lat, time_vec, origin, u10_vals)
build_synthetic_nc(v10_file, "v10", "m s-1", lon, lat, time_vec, origin, v10_vals)
build_synthetic_mask(mask_file, lon, lat)

month_frac <- c("0417", "125", "2083", "2917", "375", "4583",
                "5417", "625", "7083", "7917", "875", "9583")
build_synthetic_psmsl_station <- function(dir, station_id, years,
                                          base_level, trend_mm_per_year) {
  lines <- character(0)
  for (y in years) {
    for (m in seq_len(12)) {
      level <- base_level + trend_mm_per_year * (y - years[1]) + rnorm(1, sd = 15)
      lines <- c(lines, sprintf("%d.%s;%.1f;0;000", y, month_frac[m], level))
    }
  }
  writeLines(lines, file.path(dir, paste0(station_id, ".txt")))
}

psmsl_dir <- tempfile("psmsl_")
dir.create(psmsl_dir)
build_synthetic_psmsl_station(psmsl_dir, 1,  2011:2013, base_level = 7020, trend_mm_per_year = 3)
build_synthetic_psmsl_station(psmsl_dir, 61, 2011:2013, base_level = 6980, trend_mm_per_year = 4)

reference_period <- c("2011-01-01", "2012-12-31")  # 2 years
study_period      <- c("2011-01-01", "2013-12-31")  # 3 years
```

`reference_period` spans 2 full years rather than 1 (a single reference year
would give an undefined per-month standard deviation, and an empty
sea-level result — see `vignette("xaci-components")`), and `study_period`
extends a year beyond it. That second point matters for reading the output
below: standardising exactly 2 reference-year observations against their
own mean/sd always gives exactly `+1/sqrt(2)` or `-1/sqrt(2)` (≈ ±0.71) —
a mathematical identity, not a data-dependent result — so every 2011/2012
row is expected to look like that. 2013, being outside `reference_period`,
is not subject to this and shows genuine anomalies instead.

We use `"FRA"` as the `country_abbrev` here (rather than a fictitious code)
because the sea-level component looks up stations by matching this value
against the bundled PSMSL metadata's `Country` column — see
`vignette("xaci-components")` for details.

## Running the pipeline

With explicit `*_data_path` arguments, `calculate_aci()` skips the
ERA5-path-building step (normally driven by `years`) and goes straight to
computing every component, then combining them:

```{r, message = FALSE}
monthly_national_aci <- calculate_aci(
  country_abbrev           = "FRA",
  study_period               = study_period,
  reference_period            = reference_period,
  temperature_data_path      = t2m_file,
  precipitation_data_path    = tp_file,
  wind_u10_data_path          = u10_file,
  wind_v10_data_path          = v10_file,
  mask_data_path               = mask_file,
  sealevel_dir                  = psmsl_dir,
  granularity                    = "month",
  area                            = TRUE,
  factor                          = 1 / 5,
  admin_level                     = NULL,
  save                             = FALSE,
  computed_components              = FALSE
)

head(monthly_national_aci, 3)   # inside reference_period: components ~ +-0.71
tail(monthly_national_aci, 3)   # 2013, outside reference_period
```

## Reading the output

For national, monthly output, `calculate_aci()` returns a `data.frame` with
one row per month (`"YYYY-MM"` row names) and one column per component plus
`ACI` itself:

```{r}
colnames(monthly_national_aci)
class(monthly_national_aci)
```

Each component column is a standardised anomaly (roughly, number of
standard deviations from the reference-period mean for that calendar
month); `ACI` is their combination via the formula introduced in
`vignette("xaci-intro")`.

## Changing granularity without recomputing

Because the underlying grid-cell computation is independent from the
temporal aggregation step, you can request a different `granularity`
without paying the cost of recomputing every component — in real usage this
is best done via the `save` / `computed_components` caching pattern shown in
`vignette("xaci-components")`; here, for the small synthetic example, we
simply call `calculate_aci()` again:

```{r, message = FALSE}
seasonal_national_aci <- calculate_aci(
  country_abbrev           = "FRA",
  study_period               = study_period,
  reference_period            = reference_period,
  temperature_data_path      = t2m_file,
  precipitation_data_path    = tp_file,
  wind_u10_data_path          = u10_file,
  wind_v10_data_path          = v10_file,
  mask_data_path               = mask_file,
  sealevel_dir                  = psmsl_dir,
  granularity                    = "season",
  area                            = TRUE
)

seasonal_national_aci
```

## Grid-cell mode: output for mapping

Setting `area = FALSE` with `admin_level = NULL` switches to **grid-cell
mode**: instead of a national scalar per month, you get the full spatial
field for each component and for `ACI` itself, ready for mapping (see
`vignette("xaci-visualization")`):

```{r, message = FALSE}
grid_aci <- calculate_aci(
  country_abbrev           = "FRA",
  study_period               = study_period,
  reference_period            = reference_period,
  temperature_data_path      = t2m_file,
  precipitation_data_path    = tp_file,
  wind_u10_data_path          = u10_file,
  wind_v10_data_path          = v10_file,
  mask_data_path               = mask_file,
  sealevel_dir                  = psmsl_dir,
  granularity                    = "month",
  area                            = FALSE,
  admin_level                     = NULL,
  # See the note in vignette("xaci-components") on why this toy example
  # widens max_dist_km beyond its 500 km default.
  max_dist_km                     = 800
)

names(grid_aci)
dim(grid_aci$ACI)   # [lon x lat x time]
```

`grid_aci$lon` / `grid_aci$lat` are the coordinate vectors, `grid_aci$time`
the (aggregated) time steps, and `grid_aci$ACI`, `grid_aci$t90`, ... are
`[lon x lat x time]` arrays, each self-sufficient (carrying its own
`lon`/`lat`/`time`/`country_abbrev` attributes) once extracted from the
list.

For aggregation at the level of administrative units (e.g. French
departments) instead of nationally or on the raw grid, see
`vignette("xaci-admin-levels")`.

```{r, include = FALSE}
unlink(c(t2m_file, tp_file, u10_file, v10_file, mask_file))
unlink(psmsl_dir, recursive = TRUE)
```
