---
title: "Administrative-Level Aggregation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Administrative-Level Aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(xaci)
```

Besides the national scalar mode (`area = TRUE`) and the grid-cell mode
(`area = FALSE`, see `vignette("xaci-visualization")`), xaci can aggregate
components and the ACI itself at the level of **administrative units** —
e.g. French departments or regions — by passing `admin_level` to
`calculate_aci()` or to any individual `*_component()` function.

## How it works

Administrative aggregation relies on two pieces of geographic data, both
fetched over the network the first time they are needed:

* **`build_admin_mask()`** downloads administrative polygons via
  [GADM](https://gadm.org/) (through the `geodata` package) for the
  requested `country_abbrev` / `admin_level`, and computes, for every ERA5
  grid cell, the fraction of its area that falls inside each administrative
  unit. This is what turns a `[lon x lat x time]` array into one column per
  administrative unit.
* **`assign_sealevel_to_admin()`** does the analogous job for the sea-level
  component: it matches PSMSL tide-gauge stations to the administrative unit
  they fall in, and computes each unit's coastal fraction (the $\alpha$ in
  the ACI formula) from a world coastline layer (`rnaturalearth`).

Both of these require network access and are therefore **not run** in this
vignette. In an interactive session, the call looks like this:

```{r, eval = FALSE}
admin_mask_L1 <- build_admin_mask(
  lon = grid_aci$lon, lat = grid_aci$lat,
  country_abbrev = "FRA", admin_level = 1, crs_metric = 2154   # Lambert-93
)

dept_assignment <- assign_sealevel_to_admin(
  country_abbrev = "FRA", admin_level = 1, crs_metric = 2154
)

dept_aci_FRA <- calculate_aci(
  country_abbrev      = "FRA",
  study_period          = c("2011-01-01", "2015-12-31"),
  reference_period       = c("2011-01-01", "2013-12-31"),
  years                    = 2011:2015,
  granularity               = "month",
  admin_level                = 1,
  crs_metric                  = 2154,
  load_dir                     = "results/FRA",
  computed_components           = TRUE
)

dim(dept_aci_FRA)   # months x (7 variables x number of departments)
plot_aci_map(dept_aci_FRA, variable = "ACI", time_index = "mean")
```

Passing `admin_mask` (and, for sea level, `admin_assignment`) explicitly to
an individual component avoids rebuilding it on every call — useful when
computing several components for the same country/level:

```{r, eval = FALSE}
prec_admin_L1 <- precipitation_component(
  precipitation_data_path = "data/era5/FRA/tp_2011_2015.nc",
  country_abbrev            = "FRA",
  reference_period           = c("2011-01-01", "2013-12-31"),
  study_period                = c("2011-01-01", "2015-12-31"),
  mask_path                    = "data/era5/FRA/mask_FRA.nc",
  area                          = FALSE,
  admin_mask                    = admin_mask_L1,
  computed_components            = TRUE,
  load_dir                        = "results/FRA"
)
```

## The aggregation mechanics, without the network

The two GADM/`rnaturalearth` downloads above only *produce* two simple data
structures — an `admin_mask` (area-fraction weights per grid cell per
administrative unit) and an `admin_assignment` (station-to-unit mapping and
coastal fractions). Everything downstream of them is ordinary R and can be
demonstrated with hand-built versions of these structures, exactly as the
package's own test suite does to keep tests network-free.

### A synthetic two-region country

```{r}
lon <- c(-1, 0)
lat <- c(43, 44)

# A minimal admin_mask: 2 grid cells, cell (i, j=1) entirely in region "A",
# cell (i, j=2) entirely in region "B" (weight 1 -- no partial overlap here,
# to keep the arithmetic easy to follow).
synthetic_admin_mask <- list(
  units   = c("A", "B"),
  lon     = lon,
  lat     = lat,
  weights = list(
    `1` = c(A = 1, B = 0),   # cell (lon[1], lat[1])
    `2` = c(A = 0, B = 1),   # cell (lon[1], lat[2])
    `3` = c(A = 1, B = 0),   # cell (lon[2], lat[1])
    `4` = c(A = 0, B = 1)    # cell (lon[2], lat[2])
  )
)
```

We reuse a small synthetic temperature component (built as in
`vignette("xaci-components")`) and aggregate it per region with
`reduce_dataarray_to_dataframe()` — the same function `calculate_aci()`
calls internally once `admin_mask` is known:

```{r}
build_synthetic_t2m <- function(path, lon, lat, time_vec, origin) {
  time_hours <- as.numeric(difftime(time_vec, origin, units = "hours"))
  nlo <- length(lon); nla <- length(lat); nt <- length(time_vec)
  set.seed(5)
  trend    <- seq_len(nt) / nt   # mild warming trend, see note below
  seasonal <- 288 + 10 * sin(2 * pi * seq_len(nt) / (24 * 365)) + 0.6 * trend
  vals <- array(NA_real_, dim = c(nlo, nla, nt))
  for (i in seq_len(nlo)) {
    for (j in seq_len(nla)) {
      # A clear north/south offset so the two regions differ visibly
      vals[i, j, ] <- seasonal + 4 * j + rnorm(nt, sd = 1.5)
    }
  }
  dim_lon  <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat  <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  dim_time <- ncdf4::ncdim_def(
    "time", paste0("hours since ", format(origin, "%Y-%m-%d %H:%M:%S")),
    time_hours, unlim = TRUE
  )
  var_t2m <- ncdf4::ncvar_def("t2m", "K", list(dim_lon, dim_lat, dim_time),
                              missval = NA, prec = "double")
  nc <- ncdf4::nc_create(path, list(var_t2m))
  ncdf4::ncvar_put(nc, var_t2m, vals)
  ncdf4::nc_close(nc)
  invisible(path)
}

origin   <- as.POSIXct("1900-01-01 00:00:00", tz = "UTC")
time_vec <- seq(as.POSIXct("2011-01-01 00:00", tz = "UTC"),
                as.POSIXct("2012-12-31 23:00", tz = "UTC"), by = "hour")
t2m_file <- tempfile(fileext = ".nc")
build_synthetic_t2m(t2m_file, lon, lat, time_vec, origin)

# study_period intentionally extends a year beyond reference_period: with a
# single reference year and study_period == reference_period, every month
# would standardise to exactly 0 (compared only to itself) -- see
# vignette("xaci-intro") for why. The extra year (2012) below shows genuine
# anomalies instead, driven by the mild warming trend in the synthetic data.
reference_period <- c("2011-01-01", "2011-12-31")
study_period      <- c("2011-01-01", "2012-12-31")

t90_grid <- temperature_component(
  temperature_data_path = t2m_file,
  country_abbrev          = "XXX",
  reference_period         = reference_period,
  study_period             = study_period,
  percentile               = 90, extremum = "max", above_thresholds = TRUE,
  area                     = FALSE,     # keep the spatial dimension
  admin_mask                = synthetic_admin_mask
)

head(t90_grid, 3)   # 2011, the reference year itself: always 0
tail(t90_grid, 3)   # 2012: genuine anomalies
```

`temperature_component()` returned one column per region directly, because
we passed `admin_mask` in. Internally, each grid cell is first standardised
against its *own* reference-period distribution, and only then averaged
(weighted by area fraction) into region A and region B — so the two columns
can differ, but only insofar as the two regions' underlying cells behave
differently, not because of a raw, unstandardised temperature gap between
them.

### The sea-level equivalent

`assign_sealevel_to_admin()`'s output — `list(station_ids, factors)` — can
be built by hand the same way, using real PSMSL station IDs (see
`vignette("xaci-components")` for how to generate synthetic tide-gauge
`.txt` files for these):

```{r}
synthetic_admin_assignment <- list(
  station_ids = list(A = 1, B = 61),      # PSMSL IDs: Brest -> A, Marseille -> B
  factors     = c(A = 0.6, B = 0.3)       # coastal fraction per region
)
```

Passed as `admin_assignment` to `sealevel_component(..., area = TRUE)`, this
drives `reduce_sealevel_over_region()` to produce one `sealevel_A` /
`sealevel_B` column instead of a single national series — see
`?reduce_sealevel_over_region` for the exact aggregation logic, which mirrors
`reduce_dataarray_to_dataframe()` above but on tide-gauge station data
instead of a spatial grid.

```{r, include = FALSE}
unlink(t2m_file)
```
