---
title: "Computing the Six ACI Components"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing the Six ACI Components}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(xaci)
```

Each of the six ACI components can be computed independently of
`calculate_aci()` — useful to inspect a single component, to recompute one
after a data update, or simply to understand what each function does before
combining them. This vignette walks through all of them on a small
synthetic dataset built with the same pattern as `vignette("xaci-intro")`.

## Building a synthetic ERA5-like extract

We generate four years of hourly data (`t2m`, `tp`, `u10`, `v10`) on a tiny
3x3 grid, plus a land-mask NetCDF, entirely in memory / temp files — no
network access required. We use the first three years (2011-2013) as the
**reference period** and all four years (2011-2014) as the **study
period**: this matters for what the standardised output looks like, see the
note right after the data are generated below.

```{r}
build_synthetic_nc <- function(path, var, unit, lon, lat, time_vec, origin, vals) {
  time_hours <- as.numeric(difftime(time_vec, origin, units = "hours"))
  dim_lon  <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat  <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  dim_time <- ncdf4::ncdim_def(
    "time", paste0("hours since ", format(origin, "%Y-%m-%d %H:%M:%S")),
    time_hours, unlim = TRUE
  )
  ncvar <- ncdf4::ncvar_def(var, unit, list(dim_lon, dim_lat, dim_time),
                            missval = NA, prec = "double")
  nc <- ncdf4::nc_create(path, list(ncvar))
  ncdf4::ncvar_put(nc, ncvar, vals)
  ncdf4::nc_close(nc)
  invisible(path)
}

build_synthetic_mask <- function(path, lon, lat) {
  dim_lon <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  var_mask <- ncdf4::ncvar_def("country", "1", list(dim_lon, dim_lat),
                               missval = NA, prec = "double")
  nc <- ncdf4::nc_create(path, list(var_mask))
  ncdf4::ncvar_put(nc, var_mask, matrix(1, length(lon), length(lat)))
  ncdf4::nc_close(nc)
  invisible(path)
}

set.seed(42)
lon      <- c(-1, 0, 1)
lat      <- c(43, 44, 45)
origin   <- as.POSIXct("1900-01-01 00:00:00", tz = "UTC")
time_vec <- seq(as.POSIXct("2011-01-01 00:00", tz = "UTC"),
                as.POSIXct("2014-12-31 23:00", tz = "UTC"), by = "hour")
nlo <- length(lon); nla <- length(lat); nt <- length(time_vec)

# A mild warming/drying trend across the series, so that 2014 (the one
# out-of-reference year) shows genuine anomalies rather than pure noise.
trend <- seq_len(nt) / nt

# --- t2m: seasonal cycle in Kelvin, plus warming trend ---
seasonal_t <- 288 + 10 * sin(2 * pi * seq_len(nt) / (24 * 365)) + 0.6 * trend
t2m_vals <- array(NA_real_, c(nlo, nla, nt))
for (i in seq_len(nlo)) for (j in seq_len(nla))
  t2m_vals[i, j, ] <- seasonal_t + (i + j) + rnorm(nt, sd = 1.5)

# --- tp: precipitation (m/hour), mostly dry with occasional rain events,
# with rain becoming rarer over time (drying trend) ---
tp_vals <- array(0, c(nlo, nla, nt))
for (i in seq_len(nlo)) for (j in seq_len(nla)) {
  rain_hours <- rbinom(nt, 1, 0.08 * (1 - 0.3 * trend))
  tp_vals[i, j, ] <- rain_hours * rexp(nt, rate = 800)
}

# --- u10 / v10: wind components (m/s) ---
u10_vals <- array(rnorm(nlo * nla * nt, mean = 3, sd = 4), c(nlo, nla, nt))
v10_vals <- array(rnorm(nlo * nla * nt, mean = 1, sd = 4), c(nlo, nla, nt))

t2m_file  <- tempfile(fileext = ".nc")
tp_file   <- tempfile(fileext = ".nc")
u10_file  <- tempfile(fileext = ".nc")
v10_file  <- tempfile(fileext = ".nc")
mask_file <- tempfile(fileext = ".nc")

build_synthetic_nc(t2m_file, "t2m", "K",     lon, lat, time_vec, origin, t2m_vals)
build_synthetic_nc(tp_file,  "tp",  "m",     lon, lat, time_vec, origin, tp_vals)
build_synthetic_nc(u10_file, "u10", "m s-1", lon, lat, time_vec, origin, u10_vals)
build_synthetic_nc(v10_file, "v10", "m s-1", lon, lat, time_vec, origin, v10_vals)
build_synthetic_mask(mask_file, lon, lat)

reference_period <- c("2011-01-01", "2013-12-31")  # 3 years
study_period      <- c("2011-01-01", "2014-12-31")  # 4 years
country_abbrev    <- "XXX"   # fictitious code for this synthetic example
```

Three things worth knowing before reading the output below, all consequences
of using a small, synthetic reference period rather than the multi-decade
history you'd have with real data:

* **`reference_period` must span at least 2 years.** Standardisation divides
  by a reference-period standard deviation computed separately for each
  calendar month; with only 1 year, every month has exactly 1 reference
  sample, an undefined (`NA`) standard deviation, and (for the sea-level
  component specifically) an empty result.
* **A handful of reference years makes every component look artificially
  alike.** Standardising *n* reference observations against their own
  mean/sd always satisfies two constraints: they sum to 0, and their squares
  sum to *n* - 1. That leaves only *n* - 2 "degrees of freedom" for the
  actual data to express itself through. With `n = 2` there are **zero**
  degrees of freedom left: every month's anomaly is forced to exactly
  `+1/sqrt(2)` or `-1/sqrt(2)` (≈ ±0.71), for *every* component, regardless
  of what the underlying data actually looked like — which is why, with a
  2-year reference, temperature, precipitation, drought and wind all
  "look the same" during the reference period: they're not similar, they're
  mathematically forced into the same shape. The reference period here
  spans 3 years (1 degree of freedom per month) rather than 2 or 5+: enough
  to break that exact-magnitude artifact and let each component's
  standardised values differ from one another, while keeping this vignette
  quick to build. It's a deliberate compromise for build speed, not a fully
  realistic reference sample — with real, multi-decade data this is a
  non-issue either way.
* Months in `study_period` but *outside* `reference_period` (2014 here)
  aren't constrained this way at all and show unambiguously genuine
  anomalies, driven by the warming/drying trend built into the synthetic
  data.


## Temperature: `temperature_component()`

Both the hot-day frequency ($T_{90}$) and the cold-night frequency
($T_{10}$) come from the same function, called twice with different
`percentile` / `extremum` / `above_thresholds` arguments:

```{r}
t90 <- temperature_component(
  temperature_data_path = t2m_file,
  country_abbrev         = country_abbrev,
  reference_period        = reference_period,
  study_period            = study_period,
  mask_path               = mask_file,
  percentile              = 90, extremum = "max", above_thresholds = TRUE,
  area                    = TRUE
)

t10 <- temperature_component(
  temperature_data_path = t2m_file,
  country_abbrev         = country_abbrev,
  reference_period        = reference_period,
  study_period            = study_period,
  mask_path               = mask_file,
  percentile              = 10, extremum = "min", above_thresholds = FALSE,
  area                    = TRUE
)

head(t90, 3)   # inside reference_period (2011-2013): reflects the data
tail(t90, 3)   # 2014, outside reference_period: genuine anomalies
head(t10, 3)
tail(t10, 3)
```

Internally, `temperature_component()` averages a "day" half and a "night"
half of each day, computes the reference-period threshold at the requested
percentile, then counts (monthly) how often each day/night crosses it.

## Precipitation: `precipitation_component()`

The precipitation component is the maximum 5-day sliding sum of
precipitation, standardised against the reference period:

```{r}
prec <- precipitation_component(
  precipitation_data_path = tp_file,
  country_abbrev           = country_abbrev,
  reference_period          = reference_period,
  study_period              = study_period,
  mask_path                 = mask_file,
  var_name                  = "tp",
  window_size                = 5L,
  area                       = TRUE
)

head(prec, 3)
tail(prec, 3)
```

## Drought: `drought_component()`

The drought component is based on the maximum number of **consecutive dry
days (CDD)** per year, interpolated to monthly resolution and standardised:

```{r}
drought <- drought_component(
  precipitation_data_path = tp_file,
  country_abbrev           = country_abbrev,
  reference_period          = reference_period,
  study_period              = study_period,
  mask_path                 = mask_file,
  area                       = TRUE
)

head(drought, 3)
tail(drought, 3)
```

## Wind: `wind_component()`

The wind component measures how often wind power (derived from the `u10`
and `v10` components) exceeds its reference-period 90th percentile:

```{r}
wind <- wind_component(
  wind_u10_data_path = u10_file,
  wind_v10_data_path = v10_file,
  country_abbrev       = country_abbrev,
  reference_period      = reference_period,
  study_period          = study_period,
  mask_path             = mask_file,
  area                  = TRUE
)

head(wind, 3)
tail(wind, 3)
```

## Sea level: `sealevel_component()`

Unlike the other components, sea level comes from **PSMSL tide-gauge
records**, not gridded NetCDF data. xaci bundles the PSMSL station metadata
(names, coordinates, country) in `inst/extdata/psmsl_data.csv`, but not the
tide-gauge measurement history itself (downloaded on demand by
`request_sealevel_data()`, which requires network access).

To keep this vignette self-contained, we generate synthetic monthly
measurements for two real French stations (Brest and Marseille, PSMSL IDs 1
and 61) in the format `sealevel_load_data()` expects, instead of downloading
real records.

See the note above on why `reference_period` spans 2 full years here rather
than 1: with a single reference year, sea-level standardisation would divide
by an undefined (`NA`) per-month standard deviation, and — after that `NA`
propagates through every standardised value — return an empty result.

```{r}
psmsl_meta <- load_psmsl_data()
brest     <- psmsl_meta[psmsl_meta$ID == 1, ]
marseille <- psmsl_meta[psmsl_meta$ID == 61, ]
brest[, c("Station Name", "ID", "lat", "lon", "Country")]
marseille[, c("Station Name", "ID", "lat", "lon", "Country")]

# PSMSL encodes the month as a year fraction (see ?sealevel_correct_date_format)
month_frac <- c("0417", "125", "2083", "2917", "375", "4583",
                "5417", "625", "7083", "7917", "875", "9583")

build_synthetic_psmsl_station <- function(dir, station_id, years,
                                          base_level, trend_mm_per_year) {
  lines <- character(0)
  for (y in years) {
    for (m in seq_len(12)) {
      level <- base_level + trend_mm_per_year * (y - years[1]) + rnorm(1, sd = 15)
      lines <- c(lines, sprintf("%d.%s;%.1f;0;000", y, month_frac[m], level))
    }
  }
  writeLines(lines, file.path(dir, paste0(station_id, ".txt")))
}

psmsl_dir <- tempfile("psmsl_")
dir.create(psmsl_dir)
set.seed(123)
build_synthetic_psmsl_station(psmsl_dir, 1,  2011:2014, base_level = 7020, trend_mm_per_year = 3)
build_synthetic_psmsl_station(psmsl_dir, 61, 2011:2014, base_level = 6980, trend_mm_per_year = 4)

sealevel_national <- sealevel_component(
  country_abbrev    = "FRA",     # must match the metadata's Country column
  study_period       = study_period,
  reference_period   = reference_period,
  area                = TRUE,
  sealevel_dir        = psmsl_dir
)

head(sealevel_national, 3)   # inside reference_period (2011-2013)
tail(sealevel_national, 3)   # 2014, outside reference_period: genuine anomalies
```

`sealevel_component()` also has a grid-cell mode (`area = FALSE`), which
inverse-distance-weight-interpolates the station anomalies onto an ERA5
grid — this is what `calculate_aci()` uses internally when producing
grid-cell (mapping) output:

```{r}
sealevel_grid <- sealevel_component(
  country_abbrev  = "FRA",
  study_period     = study_period,
  reference_period = reference_period,
  area              = FALSE,
  mask_path         = mask_file,      # supplies the target lon/lat grid
  # Our fictitious grid (lon -1/0/1, lat 43/44/45) sits ~350-650 km from
  # Brest and Marseille; the default max_dist_km = 500 would leave a couple
  # of cells with no station in range (NA sea level -> NA ACI there). We
  # widen it here purely because this toy grid's placement is arbitrary --
  # with a real ERA5 grid this cutoff is meaningful and NA cells far from
  # the coast are expected, see ?sealevel_component.
  max_dist_km       = 800,
  sealevel_dir      = psmsl_dir
)

dim(sealevel_grid$data)
```

## Caching heavy computations (steps 3-4)

With real ERA5 data, each `*_component()` call above can take a long time.
Pass `save = TRUE` once to cache the result to disk, then
`computed_components = TRUE` on subsequent calls to reload it instantly —
this is exactly what `calculate_aci()` does internally:

```{r, eval = FALSE}
precipitation_component(
  precipitation_data_path = "data/era5/FRA/tp_2011_2015.nc",
  country_abbrev           = "FRA",
  reference_period          = c("2011-01-01", "2013-12-31"),
  study_period              = c("2011-01-01", "2015-12-31"),
  mask_path                 = "data/era5/FRA/mask_FRA.nc",
  area                       = FALSE,
  save                       = TRUE,
  save_dir                   = "results/FRA"
)

# Later, or in a different session:
prec_national <- precipitation_component(
  precipitation_data_path = "data/era5/FRA/tp_2011_2015.nc",
  country_abbrev           = "FRA",
  reference_period          = c("2011-01-01", "2013-12-31"),
  study_period              = c("2011-01-01", "2015-12-31"),
  mask_path                 = "data/era5/FRA/mask_FRA.nc",
  area                       = TRUE,
  computed_components        = TRUE,
  load_dir                   = "results/FRA"
)
```

The next vignette, `vignette("xaci-full-pipeline")`, shows how
`calculate_aci()` combines all six components above into the index itself.

```{r, include = FALSE}
unlink(c(t2m_file, tp_file, u10_file, v10_file, mask_file))
unlink(psmsl_dir, recursive = TRUE)
```
