---
title: "Memory-Safe Processing with the terra Engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Memory-Safe Processing with the terra Engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(xaci)
```

## Why a second engine?

The default pipeline (`temperature_component()`, `wind_component()`,
`drought_component()`, `precipitation_component()`, all built on
`load_component()` / `ncdf4`) reads the **entire** hourly NetCDF cube into
memory before doing any computation. For a single ERA5 variable over a
whole country, multiple decades, at hourly resolution, this can require
tens of gigabytes of RAM before any actual work happens.

xaci provides a memory-safe alternative built on the
[`terra`](https://rspatial.github.io/terra/) package: it reads NetCDF files
lazily (metadata only, no pixel values loaded up front), reduces hourly data
to daily resolution block by block, and writes intermediate results to disk
rather than accumulating them in RAM. Once reduced to daily resolution — a
few hundred KB to a few MB even for 40+ years of data — the rest of the
pipeline (percentile thresholds, monthly aggregation, standardisation) is
identical to the default engine: same functions, same results. This is
verified in the package's own test suite by a parity test comparing both
engines on synthetic and NetCDF-backed data.

**Rule of thumb**: use `engine = "base"` (the default) for country-level,
multi-year studies; switch to `engine = "terra"` when the default pipeline
runs out of memory, typically a whole-country, hourly, multi-decade
`study_period`.

## Same inputs, same outputs

Each memory-safe component has a `_terra`-suffixed, drop-in equivalent, used
exactly like its base-R counterpart (same arguments — plus `cores`, used
only by the temperature percentile step, see below):

| Base-R function             | `terra` equivalent                |
|------------------------------|-------------------------------------|
| `temperature_component()`    | `temperature_component_terra()`     |
| `wind_component()`           | `wind_component_terra()`            |
| `drought_component()`        | `drought_component_terra()`         |
| `precipitation_component()`  | `precipitation_component_terra()`   |

`sealevel_component()` has no `_terra` equivalent: its input is tide-gauge
CSV data, not gridded NetCDF, so the memory bottleneck this engine addresses
does not apply to it.

We build the same kind of tiny synthetic NetCDF used elsewhere in this
package's vignettes, and compute `temperature_component()` both ways to
show they agree. As in the other vignettes, `reference_period` spans 3
years and `study_period` extends a year beyond it, with a mild warming
trend added to the synthetic data — see `vignette("xaci-components")` for
why a short reference period is best avoided: with 1 year, an undefined
per-month standard deviation; with exactly 2, standardising always
collapses to exactly `+1/sqrt(2)` or `-1/sqrt(2)` (a mathematical identity,
not a data-dependent result) — informative for neither engine, and not a
meaningful check of whether they actually agree on the underlying
arithmetic. 3 reference years leave 1 "degree of freedom" per month (see
`vignette("xaci-components")` for the exact argument) -- enough to break
that shared artifact and keep this vignette quick to build, though not as
fully realistic as a longer reference sample would be.

This vignette's grid (2x2 cells, 4 years of hourly data) is deliberately
tiny so it builds quickly — it demonstrates that the `terra` engine is a
drop-in replacement with matching output, not that it holds up at country
scale over decades. That larger-scale, non-toy validation is what the
package's own test suite (`tests/testthat/test-*-terra-*.R`) is for,
comparing both engines against synthetic *and* NetCDF-backed data closer to
real ERA5 extracts.

```{r}
build_synthetic_t2m <- function(path, lon, lat, time_vec, origin) {
  time_hours <- as.numeric(difftime(time_vec, origin, units = "hours"))
  nlo <- length(lon); nla <- length(lat); nt <- length(time_vec)

  trend    <- seq_len(nt) / nt   # mild warming trend, see note above
  seasonal <- 288 + 10 * sin(2 * pi * seq_len(nt) / (24 * 365)) + 0.6 * trend
  vals <- array(NA_real_, dim = c(nlo, nla, nt))
  for (i in seq_len(nlo)) {
    for (j in seq_len(nla)) {
      vals[i, j, ] <- seasonal + (i + j) + rnorm(nt, sd = 1.5)
    }
  }

  dim_lon  <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat  <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  dim_time <- ncdf4::ncdim_def(
    "time", paste0("hours since ", format(origin, "%Y-%m-%d %H:%M:%S")),
    time_hours, unlim = TRUE
  )
  var_t2m <- ncdf4::ncvar_def("t2m", "K", list(dim_lon, dim_lat, dim_time),
                              missval = NA, prec = "double")

  nc <- ncdf4::nc_create(path, list(var_t2m))
  ncdf4::ncvar_put(nc, var_t2m, vals)
  ncdf4::nc_close(nc)
  invisible(path)
}

set.seed(99)
lon      <- c(-1, 0)
lat      <- c(43, 44)
origin   <- as.POSIXct("1900-01-01 00:00:00", tz = "UTC")
time_vec <- seq(as.POSIXct("2011-01-01 00:00", tz = "UTC"),
                as.POSIXct("2014-12-31 23:00", tz = "UTC"), by = "hour")

t2m_file <- tempfile(fileext = ".nc")
build_synthetic_t2m(t2m_file, lon, lat, time_vec, origin)

reference_period <- c("2011-01-01", "2013-12-31")  # 3 years
study_period      <- c("2011-01-01", "2014-12-31")  # 4 years
```

```{r}
t90_base <- temperature_component(
  temperature_data_path = t2m_file,
  country_abbrev          = "XXX",
  reference_period         = reference_period,
  study_period             = study_period,
  percentile               = 90, extremum = "max", above_thresholds = TRUE,
  area                     = TRUE
)

t90_terra <- temperature_component_terra(
  temperature_data_path = t2m_file,
  country_abbrev          = "XXX",
  reference_period         = reference_period,
  study_period             = study_period,
  percentile               = 90, extremum = "max", above_thresholds = TRUE,
  area                     = TRUE,
  cores                    = 1L
)

# 2011-2013 (the 3 reference years) reflect the actual data in both engines
# (see note above); 2014 carries the genuine, trend-driven signal on top --
# a much more informative comparison than a 2-year reference would give.
rbind(base = round(t90_base, 4), terra = round(t90_terra, 4))
all.equal(as.numeric(t90_base), as.numeric(t90_terra), tolerance = 1e-6)
```

On real, large data the two would differ only in memory footprint and
runtime, not in the values returned.

## Via `calculate_aci()`

Rather than swapping individual functions, pass `engine = "terra"` to
`calculate_aci()` directly — everything else about the call stays the same
(see `vignette("xaci-full-pipeline")` for the base-engine version of this
call):

```{r, eval = FALSE}
calculate_aci(
  country_abbrev      = "FRA",
  study_period         = c("2000-01-01", "2024-12-31"),
  reference_period      = c("2000-01-01", "2012-12-31"),
  years                  = 2000:2024,
  granularity             = "month",
  area                     = TRUE,
  factor                   = 1 / 5,
  admin_level               = NULL,
  save                       = TRUE,
  save_dir                   = "results/FRA",
  computed_components         = FALSE,
  engine                       = "terra",   # <- memory-safe loading & reduction
  cores                         = 12         # <- forwarded to temperature_component_terra()
)
```

`engine = "base"` (the default) reproduces the original behaviour exactly.
Cached `.rds` files are named identically regardless of which engine
produced them, so a later call with `computed_components = TRUE` reloads
them the same way either way.

## The `cores` argument

`cores` is only meaningful for `engine = "terra"`, and only affects
`temperature_component_terra()`: its rolling-percentile step
(`terra::roll()`, via `calculate_percentiles_terra()`) is by far the most
expensive part of the terra pipeline and has no native parallelisation, so
`cores` controls a manual chunked-parallel implementation instead. It is
silently ignored when `engine = "base"` (the base `temperature_component()`
has no such parameter, so `calculate_aci()` only forwards `cores` when
`engine = "terra"`).

```{r, include = FALSE}
unlink(t2m_file)
```
