---
title: "Get Started with xaci"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with xaci}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## What is the Actuarial Climate Index?

The **Actuarial Climate Index (ACI)** is a standardised measure of the
frequency and severity of extreme climate events, originally developed by
the American Academy of Actuaries and adapted here for France/Europe
following Garrido, Milhaud & Olympio (2026). It combines six components,
each expressed as a standardised anomaly relative to a reference period:

$$ACI = \frac{T_{90} - T_{10} + P + D + \alpha \cdot SL + W}{5 + \alpha}$$

| Symbol   | Component                                    |
|----------|-----------------------------------------------|
| $T_{90}$ | Frequency of hot days (90th percentile)        |
| $T_{10}$ | Frequency of cold nights (10th percentile)     |
| $P$      | Maximum sliding precipitation (5-day window)   |
| $D$      | Consecutive dry days (CDD)                     |
| $SL$     | Standardised sea level                         |
| $W$      | Wind power above the 90th percentile           |
| $\alpha$ | Coastal fraction (national default $1/5$)      |

**xaci** computes each of these components from gridded ERA5 climate data
(temperature, precipitation, wind) and from PSMSL tide-gauge records
(sea level), and aggregates them into the index above at whatever temporal
granularity (monthly, seasonal, semester, annual) and spatial level
(national, administrative unit, or grid cell) you need.

```{r setup}
library(xaci)
```

## The four-step workflow

Because ERA5 data can be large (hourly, multi-decade, whole-country), xaci
splits the work into four steps. Steps 1-3 are one-off, heavy computations;
step 4 is the fast, repeatable step used for day-to-day exploration.

```
Step 1 (long)   Step 2 (short)   Step 3 (long)              Step 4 (fast, repeatable)
ERA5 download → Country mask  → Grid-cell components     →  ACI / components at any
(data/)          (data/)         (results/, cached .rds)     temporal & spatial level
```

* **Step 1** downloads the four required ERA5 variables (`t2m`, `tp`, `u10`,
  `v10`) via `download_era5_all()` / `download_era5()` (requires a free
  Copernicus CDS account and the `ecmwfr` package).
* **Step 2** downloads a land/sea country mask with `download_mask()`.
* **Step 3** computes each component at grid-cell / monthly resolution and
  caches it to disk (`save = TRUE`) — this is the expensive step, done once
  per country and reference period.
* **Step 4** reloads the cached components (`computed_components = TRUE`)
  and re-aggregates them on the fly to any granularity or spatial level you
  ask for — this is the step you repeat while exploring the data.

`calculate_aci()` runs the whole pipeline (steps 3-4 combined, or step 4
alone if `computed_components = TRUE`); each component can also be computed
individually with `temperature_component()`, `precipitation_component()`,
`drought_component()`, `wind_component()` and `sealevel_component()`. This
is covered in detail in `vignette("xaci-components")`.

Because step 1 requires network access and a personal CDS token, it cannot
be demonstrated inside a vignette. Here is the shape of the call you would
use with your own data:

```{r, eval = FALSE}
library(ecmwfr)
cds_set_key("xxxxxxxx-xxxx-xxxx-xxxx-xxxxxxxxxxxx")

download_era5_all(
  years          = 2010:2020,
  area           = c(51.5, -5.5, 41.0, 10.0),   # N, W, S, E (metropolitan France)
  country_abbrev = "FRA"
)

download_mask(
  country_abbrev = "FRA",
  area           = c(51.5, -5.5, 41.0, 10.0),
  dest_dir       = "data/era5/FRA"
)
```

## A minimal worked example

The rest of this vignette (and the ones that follow) instead work on a tiny,
self-contained **synthetic** dataset built in a few lines of R, so that the
code below runs anywhere, with no downloads and no CDS account. The synthetic
data mimics the structure of a real ERA5 extract: an hourly NetCDF file with
`longitude` / `latitude` / `time` dimensions, plus a `country` land-mask
NetCDF, exactly what `temperature_component()` and friends expect.

```{r}
build_synthetic_t2m <- function(path, lon, lat, time_vec, origin) {
  time_hours <- as.numeric(difftime(time_vec, origin, units = "hours"))
  nlo <- length(lon); nla <- length(lat); nt <- length(time_vec)

  set.seed(1)
  seasonal <- 288 + 10 * sin(2 * pi * seq_len(nt) / (24 * 365))  # ~15C in Kelvin
  warming  <- 0.6 * seq_len(nt) / nt   # a mild warming trend across the series
  vals <- array(NA_real_, dim = c(nlo, nla, nt))
  for (i in seq_len(nlo)) {
    for (j in seq_len(nla)) {
      vals[i, j, ] <- seasonal + warming + (i + j) + rnorm(nt, sd = 1.5)
    }
  }

  dim_lon  <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat  <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  dim_time <- ncdf4::ncdim_def(
    "time", paste0("hours since ", format(origin, "%Y-%m-%d %H:%M:%S")),
    time_hours, unlim = TRUE
  )
  var_t2m <- ncdf4::ncvar_def("t2m", "K", list(dim_lon, dim_lat, dim_time),
                              missval = NA, prec = "double")

  nc <- ncdf4::nc_create(path, list(var_t2m))
  ncdf4::ncvar_put(nc, var_t2m, vals)
  ncdf4::nc_close(nc)
  invisible(path)
}

build_synthetic_mask <- function(path, lon, lat) {
  dim_lon <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  var_mask <- ncdf4::ncvar_def("country", "1", list(dim_lon, dim_lat),
                               missval = NA, prec = "double")
  nc <- ncdf4::nc_create(path, list(var_mask))
  # All cells fully "inside" the (fictitious) country for this example
  ncdf4::ncvar_put(nc, var_mask, matrix(1, length(lon), length(lat)))
  ncdf4::nc_close(nc)
  invisible(path)
}
```

We build four years of hourly data on a tiny 2x2 grid:

```{r}
lon      <- c(-1, 0)
lat      <- c(43, 44)
origin   <- as.POSIXct("1900-01-01 00:00:00", tz = "UTC")
time_vec <- seq(as.POSIXct("2011-01-01 00:00", tz = "UTC"),
                as.POSIXct("2014-12-31 23:00", tz = "UTC"), by = "hour")

t2m_file  <- tempfile(fileext = ".nc")
mask_file <- tempfile(fileext = ".nc")

build_synthetic_t2m(t2m_file, lon, lat, time_vec, origin)
build_synthetic_mask(mask_file, lon, lat)
```

...and compute the T90 component (frequency of hot days), standardised
against 2011 as a **reference** period, over the full two years as the
**study** period. This distinction matters: if `study_period` were set equal
to `reference_period` here (a single year), every month would have exactly
one observation to compare against itself, and the standardised anomaly
would be identically zero for every month, by construction — not a bug, just
a degenerate case worth being aware of when experimenting with short,
synthetic examples like this one.

```{r}
reference_period <- c("2011-01-01", "2013-12-31")
study_period      <- c("2011-01-01", "2014-12-31")

t90 <- temperature_component(
  temperature_data_path = t2m_file,
  country_abbrev         = "XXX",
  reference_period       = reference_period,
  study_period            = study_period,
  mask_path               = mask_file,
  percentile              = 90,
  extremum                = "max",
  above_thresholds        = TRUE,
  area                    = TRUE     # national (spatially averaged) series
)

head(t90, 14)
```

Note that every 2011 month is exactly `0`: that year *is* the reference
period, so each of its months is being compared to itself, by construction.
2012's months show real anomalies instead, since the mild warming trend
built into the synthetic data (see the code above) makes 2012 run warmer
than the 2011 baseline.

`t90` is a named numeric vector: one standardised value per month, positive
values indicating months with more hot days than the reference average,
negative values fewer. This is exactly the kind of object that feeds into
`calculate_aci()`. The next vignettes build on this same synthetic-data
pattern to cover:

* `vignette("xaci-components")` — computing all six components individually,
  including the sea-level component from PSMSL station data.
* `vignette("xaci-full-pipeline")` — running `calculate_aci()` end to end and
  interpreting its output.
* `vignette("xaci-visualization")` — plotting time series, component
  breakdowns, distributions and maps.
* `vignette("xaci-terra-engine")` — the memory-safe `engine = "terra"` option
  for whole-country, multi-decade, hourly historical periods.
* `vignette("xaci-admin-levels")` — aggregating results at the administrative
  unit level (e.g. French departments) instead of nationally.

```{r, include = FALSE}
unlink(c(t2m_file, mask_file))
```
