---
title: "Visualizing ACI Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing ACI Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6.5,
  fig.height = 4.2
)
```

```{r setup}
library(xaci)
```

xaci ships six plotting functions covering the two shapes of output produced
by `calculate_aci()` (see `vignette("xaci-full-pipeline")`): time-indexed
`data.frame`s (national or administrative level) and spatial grids
(grid-cell level, for mapping). This vignette demonstrates each of them on
the same synthetic dataset used throughout this package's vignettes.

## Preparing example results

```{r, echo = FALSE}
build_synthetic_nc <- function(path, var, unit, lon, lat, time_vec, origin, vals) {
  time_hours <- as.numeric(difftime(time_vec, origin, units = "hours"))
  dim_lon  <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat  <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  dim_time <- ncdf4::ncdim_def(
    "time", paste0("hours since ", format(origin, "%Y-%m-%d %H:%M:%S")),
    time_hours, unlim = TRUE
  )
  ncvar <- ncdf4::ncvar_def(var, unit, list(dim_lon, dim_lat, dim_time),
                            missval = NA, prec = "double")
  nc <- ncdf4::nc_create(path, list(ncvar))
  ncdf4::ncvar_put(nc, ncvar, vals)
  ncdf4::nc_close(nc)
  invisible(path)
}

build_synthetic_mask <- function(path, lon, lat) {
  dim_lon <- ncdf4::ncdim_def("longitude", "degrees_east", lon)
  dim_lat <- ncdf4::ncdim_def("latitude", "degrees_north", lat)
  var_mask <- ncdf4::ncvar_def("country", "1", list(dim_lon, dim_lat),
                               missval = NA, prec = "double")
  nc <- ncdf4::nc_create(path, list(var_mask))
  ncdf4::ncvar_put(nc, var_mask, matrix(1, length(lon), length(lat)))
  ncdf4::nc_close(nc)
  invisible(path)
}

set.seed(11)
lon      <- c(-1, 0, 1)
lat      <- c(43, 44, 45)
origin   <- as.POSIXct("1900-01-01 00:00:00", tz = "UTC")
time_vec <- seq(as.POSIXct("2011-01-01 00:00", tz = "UTC"),
                as.POSIXct("2013-12-31 23:00", tz = "UTC"), by = "hour")
nlo <- length(lon); nla <- length(lat); nt <- length(time_vec)

# A mild warming/drying trend, so 2013 (the one out-of-reference year, see
# vignette("xaci-components") for why reference_period stops at 2012) shows
# genuine anomalies rather than pure noise -- this is what gives the plots
# below actual visual structure instead of flat, uninformative series.
trend <- seq_len(nt) / nt

seasonal_t <- 288 + 10 * sin(2 * pi * seq_len(nt) / (24 * 365)) + 0.6 * trend
t2m_vals <- array(NA_real_, c(nlo, nla, nt))
for (i in seq_len(nlo)) for (j in seq_len(nla))
  t2m_vals[i, j, ] <- seasonal_t + (i + j) + rnorm(nt, sd = 1.5)

tp_vals <- array(0, c(nlo, nla, nt))
for (i in seq_len(nlo)) for (j in seq_len(nla)) {
  rain_hours <- rbinom(nt, 1, 0.08 * (1 - 0.3 * trend))
  tp_vals[i, j, ] <- rain_hours * rexp(nt, rate = 800)
}

u10_vals <- array(rnorm(nlo * nla * nt, mean = 3, sd = 4), c(nlo, nla, nt))
v10_vals <- array(rnorm(nlo * nla * nt, mean = 1, sd = 4), c(nlo, nla, nt))

t2m_file  <- tempfile(fileext = ".nc")
tp_file   <- tempfile(fileext = ".nc")
u10_file  <- tempfile(fileext = ".nc")
v10_file  <- tempfile(fileext = ".nc")
mask_file <- tempfile(fileext = ".nc")

build_synthetic_nc(t2m_file, "t2m", "K",     lon, lat, time_vec, origin, t2m_vals)
build_synthetic_nc(tp_file,  "tp",  "m",     lon, lat, time_vec, origin, tp_vals)
build_synthetic_nc(u10_file, "u10", "m s-1", lon, lat, time_vec, origin, u10_vals)
build_synthetic_nc(v10_file, "v10", "m s-1", lon, lat, time_vec, origin, v10_vals)
build_synthetic_mask(mask_file, lon, lat)

month_frac <- c("0417", "125", "2083", "2917", "375", "4583",
                "5417", "625", "7083", "7917", "875", "9583")
build_synthetic_psmsl_station <- function(dir, station_id, years,
                                          base_level, trend_mm_per_year) {
  lines <- character(0)
  for (y in years) {
    for (m in seq_len(12)) {
      level <- base_level + trend_mm_per_year * (y - years[1]) + rnorm(1, sd = 15)
      lines <- c(lines, sprintf("%d.%s;%.1f;0;000", y, month_frac[m], level))
    }
  }
  writeLines(lines, file.path(dir, paste0(station_id, ".txt")))
}

psmsl_dir <- tempfile("psmsl_")
dir.create(psmsl_dir)
build_synthetic_psmsl_station(psmsl_dir, 1,  2011:2013, base_level = 7020, trend_mm_per_year = 3)
build_synthetic_psmsl_station(psmsl_dir, 61, 2011:2013, base_level = 6980, trend_mm_per_year = 4)

reference_period <- c("2011-01-01", "2012-12-31")  # 2 years
study_period      <- c("2011-01-01", "2013-12-31")  # 3 years
```

We compute a national, monthly ACI series (for the time-series-style plots)
and a grid-cell version (for the map), exactly as in
`vignette("xaci-full-pipeline")`:

```{r, message = FALSE}
monthly_national_aci <- calculate_aci(
  country_abbrev           = "FRA",
  study_period               = study_period,
  reference_period            = reference_period,
  temperature_data_path      = t2m_file,
  precipitation_data_path    = tp_file,
  wind_u10_data_path          = u10_file,
  wind_v10_data_path          = v10_file,
  mask_data_path               = mask_file,
  sealevel_dir                  = psmsl_dir,
  granularity                    = "month",
  area                            = TRUE
)

grid_aci <- calculate_aci(
  country_abbrev           = "FRA",
  study_period               = study_period,
  reference_period            = reference_period,
  temperature_data_path      = t2m_file,
  precipitation_data_path    = tp_file,
  wind_u10_data_path          = u10_file,
  wind_v10_data_path          = v10_file,
  mask_data_path               = mask_file,
  sealevel_dir                  = psmsl_dir,
  granularity                    = "month",
  area                            = FALSE,
  admin_level                     = NULL,
  # See the note in vignette("xaci-components") on why this toy example
  # widens max_dist_km beyond its 500 km default -- without it, a couple of
  # cells in our fictitious grid would have no sea-level station in range
  # (NA there, and NA ACI as a result), leaving visible gaps on the map.
  max_dist_km                     = 800
)
```

## Time series: `plot_aci_timeseries()`

```{r}
plot_aci_timeseries(monthly_national_aci, smooth = TRUE, span = 0.3)
```

With only two years of monthly (24-point) synthetic data the LOESS trend
line is not very meaningful — with real, multi-decade data it highlights
the long-run direction of the index. `fill_area = TRUE` (the default) shades
the area between the series and zero, making positive/negative months
visually distinct; `colour` controls the line colour.

## Component breakdown: `plot_aci_components()`

```{r}
plot_aci_components(monthly_national_aci, type = "bar")
```

```{r}
plot_aci_components(monthly_national_aci, type = "stacked")
```

`components` restricts the plot to a subset, e.g.
`plot_aci_components(monthly_national_aci, type = "bar", components = c("t90", "t10"))`
to focus on temperature only.

## Distributions: `plot_aci_distribution()`

```{r}
plot_aci_distribution(monthly_national_aci, type = "boxplot", include_aci = TRUE)
```

```{r}
plot_aci_distribution(monthly_national_aci, type = "violin")
```

`type = "density"` is also available. This is useful to compare the spread
of each component at a glance, e.g. to spot which one is driving an
unusually high or low ACI value in a given month.

## Maps: `plot_aci_map()`

`plot_aci_map()` dispatches on its input: a grid-cell list or bare array
produces a **raster** map; a `data.frame` (administrative-level output, see
`vignette("xaci-admin-levels")`) produces a **choropleth**. Here we use the
grid-cell result computed above, averaged over the whole study period:

```{r}
plot_aci_map(grid_aci, variable = "ACI", time_index = "mean",
             borders = FALSE, title = "Mean ACI over the study period")
```

`borders = TRUE` (the default) overlays the country's administrative
boundary, downloaded on demand via `rnaturalearth` / GADM — this requires
network access, so it is disabled (`borders = FALSE`) in this
self-contained example. In an interactive session with network access, you
can drop `borders = FALSE` to get the overlay for free.

A single time slice can be plotted by passing an integer instead of
`"mean"`:

```{r}
plot_aci_map(grid_aci, variable = "t90", time_index = 6, borders = FALSE)
```

A bare array (e.g. `grid_aci$ACI` on its own, once it carries `lon`/`lat`
attributes as `calculate_aci()` attaches them) also works directly:

```{r}
plot_aci_map(grid_aci$drought, time_index = "mean", borders = FALSE,
             var_label = "Drought anomaly")
```

## Dashboard: `plot_aci_dashboard()`

`plot_aci_dashboard()` arranges the time series, component facets, boxplot
and density plots into a single figure. It requires the `patchwork`
package:

```{r}
if (requireNamespace("patchwork", quietly = TRUE)) {
  plot_aci_dashboard(monthly_national_aci)
} else {
  message("Install the 'patchwork' package to use plot_aci_dashboard().")
}
```

## Animated maps: `animate_aci_map()`

For a time-evolving view of a spatial variable, `animate_aci_map()` produces
a GIF (requires the `gganimate` and `gifski` packages). This is
computationally heavier and not run in this vignette, but the call looks
like:

```{r, eval = FALSE}
animate_aci_map(grid_aci, variable = "ACI", fps = 2,
                save_path = "aci_animation.gif")
```

```{r, include = FALSE}
unlink(c(t2m_file, tp_file, u10_file, v10_file, mask_file))
unlink(psmsl_dir, recursive = TRUE)
```
